Tropical Andes Taxonomic Diversity of Plants and Frugivores

### 2024-05-14 code/L2/L2_1_TropicalAndes_taxdiv.Rmd
update data names and paths to read in, change plot function scale color and limits, update variable names to fix errors

# Load required packages
```{r}
library(sf)
library(dplyr)
library(ggplot2)
library(rnaturalearth)
library(smoothr)
library(purrr)
library(raster)
library(ggspatial)
library(gridExtra)
```

@@

# Read in Data
```{r}
TropicalAndes_plant_occ_forest <- read.csv(file.path(data_path_L1,"TropicalAndes_GBIF_plant_occ_cleaned_subset.csv"))
TropicalAndes_frugivore_occ_forest <- read.csv(file.path(data_path_L1,"TropicalAndes_GBIF_frugivore_occ_cleaned_subset.csv"))
TropicalAndes_IUCNHabitat_Forest <- read_sf(file.path(data_path_L0, "Forest_sf.shp"), layer = "Forest_sf")
```

# Plot TD/Richness
## Function for richness plots
Need to set legend scale and color for plants constant (same for frugivores)
remove cells with richness of 0
```{r}
create_rich_plots <- function(resolution_meters) {
 # Make Grid
  TAGrid <- TApoly %>%
    st_make_grid(cellsize = c(resolution_meters)) %>%
    st_intersection(TropicalAndes_IUCNHabitat_Forest) %>%
    st_cast("MULTIPOLYGON") %>%
    st_sf() %>%
    mutate(cellid = row_number())

@@

  # Helper function to generate plots
  generate_plot <- function(data, title) {
    ggplot(data) +
      geom_sf(data = Americas, fill = "white") +
      geom_sf(data = TApoly, fill = "grey", size = 0.1) +
      geom_sf(aes(fill = num_species), color = NA) +
      scale_fill_scico(palette = "romaO", direction = 1) +
      labs(title = title, x = "Longitude", y = "Latitude", fill = "Number of species") +
      coord_sf(xlim = c(-85, -60), ylim = c(-24, 14), expand = FALSE, crs = 4326) +
      annotation_scale(location = "bl", width_hint = 0.5)  +
      theme(panel.background = element_rect(fill = "lightblue"))
  }

  # Generate plots
  plant_plot <- generate_plot(plant_richness_grid, paste("Tropical Andes Plant Richness [", resolution_meters/1000, " km]", sep=""))
  plant_plot <- generate_plot(frugivore_richness_grid, paste("Tropical Andes Frugivore Richness [", resolution_meters/1000, " km]", sep=""))

  # Extract cell values
  plant_cellRichness <- plant_plot[["data"]]
  frugivore_cellRichness <- plant_plot[["data"]]

  # Compare plant and frugivore richness
  compare_richness_summary <- summary(
    lm(frugivore_richness_grid$num_species ~ plant_richness_grid$num_species))
  r_squared <- compare_richness_summary$r.squared

  comparison_data <- data.frame(Plant_Richness = plant_richness_grid$num_species,
                               Frugivore_Richness = frugivore_richness_grid$num_species)

  ## Create scatter plot
  plant_frug_richness_compare <- ggplot(comparison_data,
                                        aes(x = Plant_Richness, y = Frugivore_Richness)) +
  geom_point() +
  labs(title = paste("Plant Richness vs Frugivore Richness [", resolution_meters/1000, " km]", sep = ""),
       subtitle =  paste("R-squared = ", round(r_squared, 2), sep = ""),
       x = "Plant Richness by cell", y = "Frugivore Richness by cell") +
  geom_smooth(method = "lm", se = FALSE, color = "red")  # Add regression line

  # Create histograms
  plant_richness_hist <- hist(plant_richness_grid$num_species, 
    main = paste("Histogram of Plant Richness [", resolution_meters/1000, " km] by cell", sep = ""),
                              xlab = "Plant Richness by cell")
  frugivore_richness_hist <- hist(frugivore_richness_grid$num_species,
    main = paste("Histogram of Frugivore Richness [", resolution_meters/1000, " km] by cell", sep = ""),
                                  xlab = "Frugivore Richness by cell")

  # Return a list of plots
  list(plantgridRichTA = plant_plot,
       frugivoregridRichTA = frugivore_plot,
       plant_cellRichness = plant_richness_grid,
       frugivore_cellRichness = frugivore_richness_grid, 
       compare_richness_summary = compare_richness_summary,
       plant_frug_richness_compare = plant_frug_richness_compare,
       plant_richness_hist = plant_richness_hist,
       frugivore_richness_hist = frugivore_richness_hist)
}

@@

# Multi panel plots
Richness maps
```{r}
all_richness_plots <- gridExtra::grid.arrange(#plantgridRichTA_1km, frugivoregridRichTA_1km,
                                              #plantgridRichTA_5km, frugivoregridRichTA_5km, 
                                               plantgridRichTA_10km, frugivoregridRichTA_10km,
                                               plantgridRichTA_25km, frugivoregridRichTA_25km,
                                               plantgridRichTA_50km, frugivoregridRichTA_50km,
                                               plantgridRichTA_75km,  frugivoregridRichTA_75km,
                                               plantgridRichTA_100km, frugivoregridRichTA_100km,
                                               ncol = 5, nrow = 2)
all_richness_plots
ggsave("all_richness_plots.pdf", all_richness_plots, path = output_path, width = 20, height = 8.5, units = "in")
```

# Compare r-squared by resolution
Plot
```{r}
# create dataframe of resolutions and r-squared values 
#[ADD 1, 5,]
resolution <- c( 10, 25, 50, 75, 100)
#ADD compare_richness_r2_1km, compare_richness_r2_5km,
r_squared <- c( compare_richness_r2_10km, compare_richness_r2_25km, compare_richness_r2_50km, compare_richness_r2_75km, compare_richness_r2_100km)
r_squared_compare <- data.frame(resolution, r_squared)

>>>>

# Load required packages
```{r}
library(sf)
library(dplyr)
library(ggplot2)
library(rnaturalearth)
library(smoothr)
library(purrr)
library(raster)
library(ggspatial)
library(ggpubr)
```

@@

# Read in Data
```{r}
## projected sf objects
plants_sf_species <- readRDS(file = file.path(data_path_L1,"plants_sf_species.rds"))
frugivores_sf_species <- readRDS(file = file.path(data_path_L1,"frugivores_sf_species.rds"))
Americas <- readRDS(file = file.path(data_path_L1, "Americas.rds"))
TApoly <- readRDS(file = file.path(data_path_L1,"TApoly.rds"))
TropicalAndes_IUCNHabitat_Forest <- readRDS(file = file.path(data_path_L1,"TropicalAndes_IUCNHabitat_Forest.rds"))
```

# Plot TD/Richness
## Function for richness plots
```{r}
create_rich_plots <- function(resolution_meters) {
 # Make Grid
  TAGrid <- TApoly %>%
    st_make_grid(cellsize = c(resolution_meters)) %>%
    st_intersection(TropicalAndes_IUCNHabitat_Forest) %>%
    st_cast("MULTIPOLYGON") %>%
    st_sf() %>%
    mutate(cellid = row_number())

@@

  # Helper functions to generate plots
  generate_plant_plot <- function(data) {
    ggplot(data) +
      geom_sf(data = Americas, fill = "white") +
      geom_sf(data = TApoly, fill = "grey", size = 0.1) +
      geom_sf(aes(fill = num_species), color = NA) +
      scale_fill_distiller(palette = "Greens", limits = c(0,6000)) +
      labs(title = title, x = "Longitude", y = "Latitude", fill = "Number of species") +
      coord_sf(xlim = c(-82, -60), ylim = c(-24, 14), expand = FALSE, crs = 4326) +
      annotation_scale(location = "bl", width_hint = 0.5)  +
      theme(panel.background = element_rect(fill = "lightblue"))
  }
  generate_frugivore_plot <- function(data) {
    ggplot(data) +
      geom_sf(data = Americas, fill = "white") +
      geom_sf(data = TApoly, fill = "grey", size = 0.1) +
      geom_sf(aes(fill = num_species), color = NA) +
      scale_fill_distiller(palette = "YlOrBr", limits = c(0,500)) + 
      labs(title = title, x = "Longitude", y = "Latitude", fill = "Number of species") +
      coord_sf(xlim = c(-82, -60), ylim = c(-24, 14), expand = FALSE, crs = 4326) +
      annotation_scale(location = "bl", width_hint = 0.5)  +
      theme(panel.background = element_rect(fill = "lightblue"))
  }

  # Generate plots
  plant_plot <- generate_plant_plot(plant_richness_grid)
  frugivore_plot <- generate_frugivore_plot(frugivore_richness_grid)

  # Extract cell values
  comparison_data <- data.frame(Plant_Richness = plant_richness_grid$num_species,
                               Frugivore_Richness = frugivore_richness_grid$num_species)

  # Filter out rows where either plant or frugivore richness is zero
  comparison_data_filtered <- comparison_data %>%
                                filter(Plant_Richness > 0 & Frugivore_Richness > 0)
  # Compare plant and frugivore richness
  ## Fit a linear model
  compare_richness_linear_summary <- summary(lm(comparison_data_filtered$Frugivore_Richness ~
                                                  comparison_data_filtered$Plant_Richness))
  linear_r_squared <- compare_richness_linear_summary$r.squared

  ## Fit an exponential model
  compare_richness_exponential_summary <- summary(lm(exp(comparison_data_filtered$Frugivore_Richness) ~
                                                       comparison_data_filtered$Plant_Richness))
  exponential_r_squared <- compare_richness_exponential_summary$r.squared

  ## Create scatter plot with linear and exponential lines
  plant_frug_richness_compare <- ggplot(comparison_data,
                                        aes(x = Plant_Richness, y = Frugivore_Richness)) +
  geom_point() +
  labs(title = paste("Plant Richness vs Frugivore Richness [", resolution_meters/1000, " km]", sep = ""),
       subtitle =  paste("Linear (Red) R-squared = ", round(linear_r_squared, 2), "\n",
                  "Exponential (Blue) R-squared = ", round(exponential_r_squared, 2)),
       x = "Plant Richness by cell", y = "Frugivore Richness by cell") +
    geom_smooth(method = "lm", se = FALSE, color = "red") +
    geom_smooth(method = "lm", se = FALSE, color = "blue", formula = y ~ exp(x), linetype = "dashed")

  # Create histograms
  plant_richness_hist <- hist(comparison_data$Plant_Richness, 
    main = paste("Histogram of Plant Richness [", resolution_meters/1000, " km] by cell", sep = ""),
                              xlab = "Plant Richness by cell")
  frugivore_richness_hist <- hist(comparison_data$Frugivore_Richness,
    main = paste("Histogram of Frugivore Richness [", resolution_meters/1000, " km] by cell", sep = ""),
                                  xlab = "Frugivore Richness by cell")
}

@@

# Multi panel plots
Richness maps
```{r}
all_richness_plots <- ggpubr::ggarrange(#plantgridRichTA_1km, plantgridRichTA_5km, 
                                              plantgridRichTA_10km, plantgridRichTA_25km,
                                              plantgridRichTA_50km, plantgridRichTA_75km,
                                              plantgridRichTA_100km,
                                              #frugivoregridRichTA_1km, frugivoregridRichTA_5km, 
                                              frugivoregridRichTA_10km, frugivoregridRichTA_25km,
                                              frugivoregridRichTA_50km, frugivoregridRichTA_75km,
                                              frugivoregridRichTA_100km, 
                                              ncol = 5, nrow = 2,
                                              labels = c("10km", "25km", ))
all_richness_plots
ggsave("all_richness_plots.pdf", all_richness_plots, path = output_path, width = 20, height = 8.5, units = "in")
```

# Compare r-squared by resolution
Plot
```{r}
# create dataframe of resolutions and r-squared values 
#[ADD 1, 5,]
resolution <- c( 10, 25, 50, 75, 100)
#ADD compare_richness_r2_1km, compare_richness_r2_5km,
r_squared <- c(compare_richness_r2_10km, compare_richness_r2_25km, compare_richness_r2_50km, compare_richness_r2_75km, compare_richness_r2_100km)
r_squared_compare <- data.frame(resolution, r_squared)
